European or American? A re-examination of the transatlantic influence over the IMF

### 2025-07-22 manuscript/manuscript.qmd
remove mearsheimer para from Hu Tong's para
Add stone reference of consensus-based decision making to intro

Qualitative approaches to this issue are often hampered by the Fund’s closed-door, consensus-based decision-making—precisely the kind of opaque process exemplified by the “gentlemen’s agreement.” On the one hand, the prevalence of informal conventions limits the utility of formal institutional analyses (e.g., voting power or veto rights). On the other, the secretive nature of negotiations means scholars often observe the outcomes, but not the underlying power struggles. While case studies such as Mexico’s bailout or the European debt crisis [@lipscy2018; @copelovitch2010a; @hodson2014] provide compelling narratives, they lack transatlantic comparability and generalisability.

@@

As an international organization formally governed by a system of weighted voting and technocratic procedures, IMF is viewed as an “accountant” [@lutz2019], primarily concerned with the internal consistency of fiscal projections to guarantee timely repayment of its financial assistance. However, empirical studies suggested that countries with low international reserves, high current account deficits, and low GDP growth are more likely to receive IMF funding assistance [@barro2005; @knight1997], despite such indicators often reflecting a limited capacity to repay. This paradox leads to a widely argued statement, particularly among realists, that international institutions are “basically a reflection of the distribution of power in the world” [@mearsheimer1994]. Generally, all studies on IMF lending outcomes agree that major power impose their influence in the process. For example, studies have shown that countries strategically important to the United States are more likely to receive IMF loans, or to receive larger amounts. Specifically, this applies to countries that align with the U.S. in United Nations General Assembly and United Nations Security Council voting [@lipscy2018; @thacker1999; @dreher2009], are in geopolitical vital regions \[\], or hold significant debts to American commercial banks [@oatley2004]. This is not surprising, as the United States, as the most powerful state, has long projected its influence over international financial institutions—much like it has done with the World Bank [@wade2002; @woods2003; @clark2020]. The question this study brings about is, how this reflection of power is distributed within a given international organization, especially one that is widely perceived to be dominated by a few major powers.

>>>>

Qualitative approaches to this issue are often hampered by the Fund’s closed-door, consensus-based decision-making [@stone2011, pp.55-56]—precisely the kind of opaque process exemplified by the “gentlemen’s agreement.” On the one hand, the prevalence of informal conventions limits the utility of formal institutional analyses (e.g., voting power or veto rights). On the other, the secretive nature of negotiations means scholars often observe the outcomes, but not the underlying power struggles. While case studies such as Mexico’s bailout or the European debt crisis [@lipscy2018; @copelovitch2010a; @hodson2014] provide compelling narratives, they lack transatlantic comparability and generalisability.

@@

As an international organization formally governed by a system of weighted voting and technocratic procedures, IMF is viewed as an “accountant” [@lutz2019], primarily concerned with the internal consistency of fiscal projections to guarantee timely repayment of its financial assistance. However, empirical studies suggested that countries with low international reserves, high current account deficits, and low GDP growth are more likely to receive IMF funding assistance [@barro2005; @knight1997], despite such indicators often reflecting a limited capacity to repay. Generally, all studies on IMF lending outcomes agree that major power impose their influence in the process. For example, studies have shown that countries strategically important to the United States are more likely to receive IMF loans, or to receive larger amounts. Specifically, this applies to countries that align with the U.S. in United Nations General Assembly and United Nations Security Council voting [@lipscy2018; @thacker1999; @dreher2009], are in geopolitical vital regions \[\], or hold significant debts to American commercial banks [@oatley2004]. This is not surprising, as the United States, as the most powerful state, has long projected its influence over international financial institutions—much like it has done with the World Bank [@wade2002; @woods2003; @clark2020]. The question this study brings about is, how this reflection of power is distributed within a given international organization, especially one that is widely perceived to be dominated by a few major powers.
